---
title: "Data Manipulation"
output: github_document
---

```{r}
library(tidyverse)
```

Let's import the datasets.

```{r}
litters_df =
  read_csv("data/FAS_litters.csv", na = c("NA", ".", ""))

litters_df = 
  janitor::clean_names(litters_df)

pups_df =
  read_csv("data/FAS_pups.csv", skip = 3, na = c("NA", ".", ""))

pups_df = 
  janitor::clean_names(pups_df)
```


## `select`

`select` selects variables from my dataframe

select specific columns

```{r}
select(litters_df, group, litter_number, gd0_weight)
```

select a range

```{r}
select(litters_df, group:gd_of_birth)
```

select by removal 

```{r}
select(litters_df, -group)
select(litters_df, -(group:gd_of_birth))
```

select by prefix 

```{r}
select(litters_df, group, starts_with("gd"))
```

rearrange by selecting 

```{r}
select(litters_df, starts_with("gd"), group)

select(litters_df, litter_number, everything())

relocate(litters_df, litter_number)
```

rename by selecting

```{r}
select(litters_df, GROUP = group, everything())

rename(litters_df, GROUP = group)
```

What if I only selet on variable?

```{r}
select(litters_df, group)

pull(litters_df, group)
```

Learning Assessment: In the pups data, select the columns containing litter number, sex, and PD ears

```{r}
select(pups_df, litter_number:pd_ears)
```


## `filter`

use `filter` to filter out rows

```{r}
filter(litters_df, gd_of_birth == 20, pups_born_alive == 6)
```

look for more than five pups alive

```{r}
filter(litters_df, pups_born_alive > 5)
filter(litters_df, pups_born_alive >= 5)
```

match to a set

```{r}
filter(litters_df, group %in% c("Con7", "Con8"))
```

filter out missing row with `drop_na()`

```{r}
drop_na(litters_df) #drop all the rows having missing values
drop_na(litters_df, gd0_weight)
```

Learning Assessment: In the pups data:

Filter to include only pups with sex 1
```{r}
filter(pups_df, sex == 1)
```

Filter to include only pups with PD walk less than 11 and sex 2
```{r}
filter(pups_df, pd_walk < 11, sex == 2)
```

## `mutate`

use `mutate()` to create or modify variables.

```{r}
mutate(
  litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group) #make variables of group column to lower case
   )
```


## `arrange`

`arrange` arranges things

```{r}
arrange(litters_df, desc(pups_born_alive)) #descending 
arrange(litters_df, group, pups_born_alive)
```


## PIPES

Let's see a bad way to do things

```{r}
litters_df =
  read_csv("data/FAS_litters.csv", na = c("NA", ".", ""))

litters_df = 
  janitor::clean_names(litters_df)

litters_df = 
  select(litters_df, group, litter_number, starts_with("gd"))

litters_df = 
  drop_na(litters_df)
```

Use pipes to avoid all this

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() |>  
  select(group, litter_number, starts_with("gd")) |>  
  drop_na() |>  
  mutate(
    wt_gain = gd18_weight - gd0_weight
  )
```


Learning Assessment: Write a chain of commands that:

loads the pups data
cleans the variable names
filters the data to include only pups with sex 1
removes the PD ears variable
creates a variable that indicates whether PD pivot is 7 or more days

```{r}
pups_df = 
  read_csv("data/FAS_pups.csv", skip = 3, na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  filter(sex == 1) |> 
  select(-pd_ears) |> 
  mutate(
    pd_7 = pd_pivot >= 7 
  )
```

